---
title: |
    | DATA 612 Project 1
    | Global Baseline Predictors and RMSE
author: "Mohamed Thasleem, Kalikul Zaman"
date: "Jun 11, 2020"
output: 
  html_document:
    df_print: paged
    theme: united
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    code_folding: hide
---

## Project 1

<li>Briefly describe the recommender system that you’re going to build out from a businessperspective, e.g. “This system recommends data science books to readers.”</li>
<li>Find a dataset, or build out your own toy dataset. As a minimum requirement for complexity,please include numeric ratings for at least five users, across at least five items, with some missingdata.
</li><li>Load your data into (for example) an R or pandas dataframe, a Python dictionary or list of lists, (oranother data structure of your choosing). From there, create a user-item matrix.
</li><li>If you choose to work with a large dataset, you’re encouraged to also create a small, relativelydense “user-item” matrix as a subset so that you can hand-verify your calculations.
</li><li>Break your ratings into separate training and test datasets.
</li><li>Using your training data, calculate the raw average (mean) rating for every user-item combination.
</li><li>Calculate the RMSE for raw average for both your training data and your test data.
</li><li>Using your training data, calculate the bias for each user and each item.
</li><li>From the raw average, and the appropriate user and item biases, calculate the baseline predictorsfor every user-item combination.
</li><li>Calculate the RMSE for the baseline predictors for both your training data and your test data.
</li>Summarize your results.


```{r project1, message=FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(kableExtra)
library(knitr)
```

### Data Processing

Built a sample dataset with Users and Books, 10 on each and randomly assigned the values with some missing values NA

```{r , message=FALSE, warning=FALSE}

# random sample of 100 ratings
set.seed(612)
df <- matrix(sample(1:5, 100, replace = TRUE), nrow = 10)

# sample dataset for splitting
split_df <- sample(1:length(df), 10, replace = FALSE)

# split the data into train_dfing dataset
train_df <- df
train_df[split_df] <- NA

# split the data into train_dfing dataset
test_df <- df
test_df[-split_df] <- NA

# create some missing values for both dataset
set.seed(612)
missing_df <- sample(1:length(df), 10, replace = FALSE)
df[missing_df] <- NA
train_df[missing_df] <- NA
test_df[missing_df] <- NA

# name of the books
users <- c("User_1","User_2","User_3","User_4","User_5","User_6","User_7","User_8","User_9","User_10")
rownames(df) <- users
rownames(train_df) <- users
rownames(test_df) <- users

# name of the users
colname <- c("Book_1","Book_2","Book_3","Book_4","Book_5","Book_6","Book_7","Book_8","Book_9","Book_10")
colnames(df) <- colname
colnames(train_df) <- colname
colnames(test_df) <- colname

# print the matrix
kable(df,caption = "User-Book Ratings") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")

```

### Train Dataset

Building a training dataset

```{r , message=FALSE, warning=FALSE}
kable(train_df,caption = "train_dfing Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### Test Dataset

Builsing a test dataset

```{r , message=FALSE, warning=FALSE}
kable(test_df,caption = "test_df Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### User-Item Matrix

Used replicate function to create an user-item matrix and finding the average

```{r , message=FALSE, warning=FALSE}

# raw average
raw_avg <- round(mean(train_df, na.rm = TRUE), 2)

# user-item matrix for raw avearge
user_item <- matrix(replicate(100, raw_avg), 10)
rownames(user_item) <- rownames(train_df)
colnames(user_item) <- colnames(train_df)

kable(user_item,caption = "User-Item Matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```


### User Bias

Calculating bias for each user using rowMeans function

```{r , message=FALSE, warning=FALSE}
# bias for each user
user_bias <- round((rowMeans(train_df, na.rm = TRUE) - raw_avg), 2)

kable(user_bias,caption = "User Bias") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### Item Bias

Calculating bias for each item using colMeans function

```{r , message=FALSE, warning=FALSE}
# bias for each item
item_bias <- round((colMeans(train_df, na.rm = TRUE) - raw_avg), 2)

kable(item_bias,caption = "Item Bias") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### Baseline Predictor

Calculating baseline predictors for every user-item combination

```{r , message=FALSE, warning=FALSE}
# calculate every user-item biases combination
com <- apply(expand.grid((as_tibble(user_bias))[[1]], (as_tibble(item_bias))[[1]]), 1, sum)

# baseline predictors for every user-item combination
baseline <- (replicate(100, raw_avg) + com)
baseline <- matrix(baseline, 10)
rownames(baseline) <- rownames(train_df)
colnames(baseline) <- colnames(train_df)

kable(baseline,caption = "Item Bias") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### RMSE Calculation

Calculating RMSE for baseline predictors for training and testing data

round((sqrt(mean((x - y)^2, na.rm = TRUE))), 2)

```{r , message=FALSE, warning=FALSE}
# function to calculate RMSE
rmse <- function(x, y) {
  round((sqrt(mean((x - y)^2, na.rm = TRUE))), 2)
}

# rmse for train_df dataset
rmse1 <- rmse(train_df, raw_avg)

# rmse for test_df dataset
rmse2 <- rmse(test_df, raw_avg)

# rmse for baseline predictors
rmse3 <- rmse(test_df, baseline)
rmse4 <- rmse(train_df, baseline)
```

### Summary

```{r , message=FALSE, warning=FALSE}
# summary of the result
kable(cbind(rmse1, rmse2, rmse3, rmse4), col.names = rep(c("Train", "Test"), 2),caption = "Summary") %>%
  add_header_above(c("Raw Average" = 2, "Baseline Predictor" = 2)) %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") 
```

