---
title: |
    | DATA 612 Project 2
    | Content-Based and Collaborative Filtering
author: "Mohamed Thasleem, Kalikul Zaman"
date: "Jun 18, 2020"
output: 
  html_document:
    df_print: paged
    theme: united
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

## Project 2

This project is to perform analysis Content-Based and Collaborative Filtering on existing dataset of user-item ratings, I have picked MovieLens dataset for this analysis


```{r project1, message=FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(kableExtra)
library(knitr)
library(recommenderlab)
library(ggplot2)
library(DT)
library(pander)
library(Matrix)
```

### Data Exploration

MovieLense Data is chosen to recommend the new / untried movies to users. We explore the the different algorithms IBCF and UBCF with Cosine similarity and compare for accuracy for evaluation.

```{r , message=FALSE, warning=FALSE}
data(MovieLense)
MovieLense
## visualize part of the matrix
image(MovieLense[1:100,1:100])

par(mfrow=c(1,2))
## number of ratings per user
hist(rowCounts(MovieLense))
## number of ratings per movie
hist(colCounts(MovieLense))

## mean rating (averaged over users)
mean(rowMeans(MovieLense))
## available movie meta information
moviemeta <- MovieLenseMeta
kable(moviemeta,caption = "MovieLens Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```



The dataset is huge and for this analysis, we have taken into consideration of those users who have rated at least 20 movies and those movies that are rated b atleast 50 users, Which will enable to perform some healthy baseline of recommendation.

```{r , message=FALSE, warning=FALSE}
movielense <- MovieLense
# Extracting data tha comprises of at least 20 ratings per user and 50 ratings
movielenseorig <- movielense
movielense <- movielense[rowCounts(movielense) > 20, colCounts(movielense) > 50]
minrowcnt <- min(rowCounts(movielense))
nrow(movielense)
ncol(movielense)
```

### Train and Test Dataset

Partition the training and test dataset on 80-20 ratio.

```{r , message=FALSE, warning=FALSE}
set.seed(612)
which_train <- sample(x = c(TRUE, FALSE), size = nrow(movielense), replace = TRUE, 
    prob = c(0.8, 0.2))

recc_data_train <- movielense[which_train, ]
recc_data_test <- movielense[!which_train, ]
```

### Building Model 

We have taken Item-Based Collaborative Filtering (ICBF) and User-Based Collaborative Filtering (UBCF) for analysis

### Item-Based Collaborative Filtering (ICBF)

Item Based collaborative filtering algorithm is applied with Cosine similarity to identify 25 neighbouring items wiht similar genre profile and base recommendations on that basis

```{r , message=FALSE, warning=FALSE}
# Find top 10 recomm movies with Item based collab filter
recc_model1 <- Recommender(data = recc_data_train, method = "IBCF", parameter = list(k = 25, 
    method = "Cosine"))
recc_model1

# Applying model to test
num_rec <- 10  # Lets recommend top 5 movies to each of users

recc_predicted1 <- predict(object = recc_model1, newdata = recc_data_test, n = num_rec)
recc_predicted1

recdf <- data.frame(user = sort(rep(1:length(recc_predicted1@items), recc_predicted1@n)), 
    rating = unlist(recc_predicted1@ratings), index = unlist(recc_predicted1@items))
```

<mark>Recommendation from ICBF Model</mark>

```{r , message=FALSE, warning=FALSE}
recdf$title <- recc_predicted1@itemLabels[recdf$index]
recdf$year <- moviemeta$year[recdf$index]
#recdf <- recdf %>% group_by(user) %>% top_n(5, recdf$rating

kable(recdf[recdf$user %in% (1:10), ],caption = "Recommendation from ICBF Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```


mark>Recommendation from  ICBF Model with Period Context Added</mark>

```{r , message=FALSE, warning=FALSE}
recdfnew <- recdf[with(recdf, order(recdf$user, -recdf$year, -round(recdf$rating))), 
    c(1, 2, 5, 4)]
#recdfnew <- recdfnew %>% group_by(user) %>% top_n(5, recdfnew$year)
kable(recdfnew[recdfnew$user %in% (1:10), ],caption = "Recommendation from  ICBF MOdel with Period Context") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### User-Based Collaborative Filtering (UBCF)

UserBased collaborative filtering algorithm is applied with Cosine similarity to identify 25 neiighbouring users wiht similar profile and base recommendations on that basis

```{r , message=FALSE, warning=FALSE}

# Find top 10 recomm movies with Item based collab filter
recc_model2 <- Recommender(data = recc_data_train, method = "UBCF", parameter = list(k = 25, 
    method = "Cosine"))
recc_model2

# Applying model to test
num_rec <- 10  # Lets recommend top 5 movies to each of users

recc_predicted2 <- predict(object = recc_model2, newdata = recc_data_test, n = num_rec)
recc_predicted2

recdfub <- data.frame(user = sort(rep(1:length(recc_predicted2@items), recc_predicted2@n)), 
    rating = unlist(recc_predicted2@ratings), index = unlist(recc_predicted2@items))
```

<mark>Recommendation from UBCF Model</mark>

```{r , message=FALSE, warning=FALSE}
recdfub$title <- recc_predicted2@itemLabels[recdfub$index]
recdfub$year <- moviemeta$year[recdfub$index]
#recdfub <- recdfub %>% group_by(user) %>% top_n(5, recdfub$rating)
kable(recdfub[recdfub$user %in% (1:10), ],caption = "Recommendation from UBCF Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```


<mark>Recommendation from UBCF Model with Period Context Added</mark>

```{r , message=FALSE, warning=FALSE}
recdfubnew <- recdfub[with(recdfub, order(recdfub$user, -recdfub$year, -round(recdfub$rating))), 
    c(1, 2, 5, 4)]
#recdfubnew <- recdfubnew %>% group_by(user) %>% top_n(5, recdfubnew$year)
kable(recdfubnew[recdfubnew$user %in% (1:10), ],caption = "Recommendation from UBCF Model with Period Context Added") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = T, color = "white", background = "#ea7872") %>%
    scroll_box(width = "100%", height = "300px")
```

### Model Comparison and validation

```{r , message=FALSE, warning=FALSE}
# Since minrowcnt is 20 , we keep the itemto keep at 15
set.seed(612)

n_fold <- 10  # k value for k fold cross validation
items_to_keep <- 15  # Items to consider in training set (less than min no of ratings )
rating_threshold <- 3.5  # Considering a rating of 3.5 as good rating across all movies

eval_sets <- evaluationScheme(data = movielense, method = "cross-validation", k = n_fold, 
    given = items_to_keep, goodRating = rating_threshold)

eval_sets

evaltrain <- getData(eval_sets, "train")  # training set
evalknown <- getData(eval_sets, "known")  # known test set
evalunknown <- getData(eval_sets, "unknown")  # unknown test set

# First, let's prepare the data for validation, as shown in the previous section.
# Since #the k-fold is the most accurate approach, we will use it here:


model_to_evaluate <- "IBCF"
model_parameters <- list(method = "Cosine")
model1_IBCF_cosine <- Recommender(data = evaltrain, method = model_to_evaluate, parameter = model_parameters)

items_to_recommend <- 10

model1_prediction <- predict(object = model1_IBCF_cosine, newdata = evalknown, n = items_to_recommend, 
    type = "ratings")
model1_predtop <- predict(object = model1_IBCF_cosine, newdata = evalknown, n = items_to_recommend, 
    type = "topNList")
model1_accuracy <- calcPredictionAccuracy(x = model1_prediction, data = evalunknown, 
    byUser = FALSE)
model1_accuracy

model_to_evaluate <- "UBCF"
model_parameters <- list(method = "cosine")
model3_UBCF_cosine <- Recommender(data = evaltrain, method = model_to_evaluate, parameter = model_parameters)

items_to_recommend <- 10

model3_prediction <- predict(object = model3_UBCF_cosine, newdata = evalknown, n = items_to_recommend, 
    type = "ratings")
model3_predtop <- predict(object = model3_UBCF_cosine, newdata = evalknown, n = items_to_recommend, 
    type = "topNList")
model3_accuracy <- calcPredictionAccuracy(x = model3_prediction, data = evalunknown, 
    byUser = FALSE)  # byUser =FALSE for model level performance metrics
model3_accuracy

# Evaluating different models, we can define a list with them We add random and
# popular to the model methods of evaluation in this comparison
models_to_evaluate <- list(IBCF_cos = list(name = "IBCF", param = list(method = "cosine")), 
    UBCF_cos = list(name = "UBCF", param = list(method = "cosine")))

# In order to evaluate the models properly, we need to test them, varying the
# number of flavors , as follows
n_recommendations <- c(1, 3, 5, 7, 10, 12, 15)

list_results <- evaluate(x = eval_sets, method = models_to_evaluate, n = n_recommendations)

plot(list_results, annotate = 1, legend = "topleft")
title("ROC curve")

plot(list_results, "prec/rec", annotate = 1, legend = "bottomright")
title("Precision-recall")

listerror <- evaluate(x = eval_sets, method = models_to_evaluate, type = "ratings")
modelcomp <- as.data.frame(sapply(avg(listerror), rbind))
modelcompnew <- as.data.frame(t(as.matrix(modelcomp)))
colnames(modelcompnew) <- c("RMSE", "MSE", "MAE")
pander(modelcompnew, caption = "Model Comparison Based On Varying Recommendation")
```


